---
title: "Basic usage of edfinr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basic usage of edfinr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE,
  eval = TRUE
)
```

## Introduction

The `edfinr` package provides a simple, consistent interface for accessing comprehensive education finance data for U.S. school districts. This vignette will help you get started with the package's core functionality.

```{r setup}
library(edfinr)
library(dplyr)
```

## Core Function: get_finance_data()

The primary function in `edfinr` is `get_finance_data()`, which provides access to comprehensive school finance data from school years 2011-12 through 2021-22. The function combines data from multiple sources:

- **Financial data**: Revenue and expenditure data from NCES F-33 survey
- **Enrollment**: Student counts from NCES Common Core of Data
- **Demographics**: Poverty estimates from Census SAIPE
- **Community characteristics**: Income and education data from American Community Survey
- **Inflation adjustments**: CPI-U data for constant dollar calculations

## Basic Usage

The simplest way to use `get_finance_data()` is to specify a year and state. For example, to get finance data for Kentucky school districts from the 2015-16 school year:
```{r example-1}
ky_sy16 <- get_finance_data(yr = 2016, geo = "KY")

# View the structure of the returned data
glimpse(ky_sy16)
```

## Dataset Types: Skinny vs. Full

By default, `get_finance_data()` returns a "skinny" dataset with 41 essential variables covering:
- District identifiers and characteristics
- Total revenues by source (local, state, federal)
- Current expenditures
- Key demographic and economic indicators

For more detailed analysis, you can request the "full" dataset with 89 variables that includes:
- All skinny dataset variables
- Detailed expenditure breakdowns by function and object
- Capital outlay and debt service details
- COVID-19 federal assistance expenditures (2020-2022)

```{r example-2}
# Get the full dataset with detailed expenditure data
ky_full_sy16 <- get_finance_data(yr = "2016", geo = "KY", dataset_type = "full")

# See what additional variables are available
names(ky_full_sy16)[42:89]
```

## Multiple Years and States

The `get_finance_data()` function makes it easy to access data across multiple years and states:

```{r example-3}
# Get data for multiple states (SEC conference states) across multiple years
sec_data <- get_finance_data(
  yr = "2018:2022",  # Years 2018 through 2022
  geo = "AL,AR,FL,GA,KY,LA,MS,MO,OK,SC,TN,TX"  # Comma-separated state codes
)

# Or get all available years for a single state
ky_all_years <- get_finance_data(yr = "all", geo = "KY")

# Or get the most recent year for all states
us_sy22 <- get_finance_data(yr = 2022, geo = "all")
```


## Working with the Data

Once you've retrieved the data, you can use standard data manipulation tools to analyze it. Here are some common analysis patterns:

### Finding High and Low Revenue Districts

```{r analysis-1}
# identify the highest and lowest revenue districts in Kentucky
ky_sy22 <- get_finance_data(yr = "2022", geo = "KY")

# districts with highest per-pupil revenue
hi_rev_dist <- ky_sy22 |>
  select(dist_name, enroll, rev_total_pp, exp_cur_pp) |>
  arrange(desc(rev_total_pp)) |>
  head(10)

print(hi_rev_dist)

# districts with lowest per-pupil revenue
lo_rev_dist <- ky_sy22 |>
  filter(enroll >= 100) |>  # focus on districts with meaningful enrollment
  select(dist_name, enroll, rev_total_pp, exp_cur_pp) |>
  arrange(rev_total_pp) |>
  head(10)

print(hi_rev_dist)
```

### Analyzing Revenue Sources

```{r analysis-2}
# compare revenue sources across districts
revenue_analysis <- ky_sy22 |>
  mutate(
    pct_local = rev_local / rev_total,
    pct_state = rev_state / rev_total,
    pct_federal = rev_fed / rev_total
  ) |>
  select(dist_name, urbanicity, pct_local, pct_state, pct_federal) |>
  group_by(urbanicity) |>
  summarize(
    avg_pct_local = mean(pct_local, na.rm = TRUE),
    avg_pct_state = mean(pct_state, na.rm = TRUE),
    avg_pct_federal = mean(pct_federal, na.rm = TRUE),
    n_districts = n()
  )

print(revenue_analysis)
```

### Working with Detailed Expenditure Data

```{r analysis-3}
# analyze COVID-19 federal assistance spending (requires full dataset)
ky_full_sy22 <- get_finance_data(yr = "2022", geo = "KY", dataset_type = "full")

covid_spending <- ky_full_sy22 |>
  filter(!is.na(exp_covid_total), exp_covid_total > 0) |>
  mutate(
    covid_pp = exp_covid_total / enroll,
    pct_of_total_exp = exp_covid_total / exp_cur_total
  ) |>
  select(dist_name, enroll, exp_covid_total, covid_pp, pct_of_total_exp) |>
  arrange(desc(covid_pp))

print(head(covid_spending, 10))
```

## Additional Resources

For more information about the data and methods used in this package:

- Use `list_variables()` to see all available variables and their descriptions
- Use `get_states()` to see valid state codes
- See the "CPI Adjustments" vignette for information about inflation adjustments
- See the "Data Sources and Methods" vignette for detailed methodology