---
title: "Basic usage of edfinr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basic usage of edfinr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

To use the `edfinr` package to access district-level school finance data, start by loading the package.

```{r setup}
library(edfinr)
```

The primary function in `edfinr` is `get_finance_data`, which allows users to download school finance data from school years 2011-12 through 2021-22. District-level revenue, expenditure, and enrollment data come from the National Center for Education Statistics (NCES) F-33 survey. Additional data are drawn from the NCES Common Core of Data Directory file (via the Urban Institute's `educationdata` package), along with the U.S. Census Bureau's 5-Year American Community Survey (via the `tidycensus` package) and Small Area Income and Poverty Estimates (SAIPE). Inflation is indexed using the U.S. Bureau of Labor Statistics Consumer Price Index for All Urban Consumers (CPI-U). 

Basic usage of `get_finance_data` starts with specifying the years and geographies of data you'd like to explore. Here's a simple example that will return finance data for Kentucky school districts from the 2015-16 school year:
```{r example-1}
ky_sy16 <- get_finance_data(yr = 2016, geo = "KY")
```

By default, `get_finance_data` returns a tibble with 41 columns. More detailed expenditure data can be added by adjusting the `dataset_type` parameter from its default (`"skinny"`) to `"full"`. This will return a tibble with 48 additional expenditure variables for a total of 89 columns.

```{r example-2}
ky_full_sy16 <- get_finance_data(yr = "2016", geo = "KY", dataset_type = "full")

```

It's straighforward to access data from multiple years and/or states with `get_finance_data`. This example will return all available years of data for each state with a university in the Southeastern Conference:

```{r example-3}
sec_sy12_thru_sy22 <- get_finance_data(yr = "2016:2022", geo = "AL,AR,FL,GA,KY,LA,MS,MO,OK,SC,TN,TX")
```

All dollar-denominated data are *nominal* by default, meaning they are *unadjusted for inflation.* Users can automatically adjust all dollar-denominated data (revenues, expenditures, median household income, and median property value) by setting the `cpi_adj` parameter (which defaults to `"none"`) to a base school year. For example, setting `cpi_adj = 2022` will adjust all dollar-denominated values for inflation using the CPI-U index for the 2021-22 school year as a baseline. 

The CPI-U for each school year is calculated by averaging the CPI-U index from "HALF2" of the first calendar year of a school year and "HALF1" of the second calendar year in a school year. For example, the 2021-22 CPI-U index is calculated by averaging "HALF2" from 2021 with "HALF1" from 2022. By default, data returned by `get_finance_data` includes a column, `cpi_sy12', that expresses the CPI-U index for each datapoint relative to the index from the 2011-12 school year (`cpi_sy12` = 1 for the 2011-12 school year). 