---
title: "Capital and Facilities"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Capital and Facilities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  eval = TRUE
)
```

## Introduction

Beyond current operating spending, the F-33 survey reports how districts fund
and build their facilities. `edfinr` exposes these as a set of capital, debt,
and fund-balance variables aimed at analysts working on school facilities.

```{r setup, include = FALSE}
library(edfinr)
library(dplyr)
library(ggplot2)
```

```{r, eval = FALSE}
library(edfinr)
library(dplyr)
library(ggplot2)
```

The skinny dataset includes total capital outlay in dollars and per-pupil terms
(`exp_cap_total`, `exp_cap_total_pp`). The **full** dataset adds the detailed
components and the debt/fund-balance picture:

- **Capital outlay components:** `exp_cap_construction`, `exp_cap_land`,
  `exp_cap_equip_instr`, `exp_cap_equip_other`, `exp_cap_equip_nonspec`.
- **Debt service:** `exp_debt_interest` (interest only).
- **Debt stocks:** `debt_lt_begin`, `debt_lt_issued`, `debt_lt_retired`,
  `debt_lt_end`, `debt_st_begin`, `debt_st_end`.
- **Fund balances (fiscal year end):** `fund_bal_debt_svc`, `fund_bal_bond`,
  `fund_bal_other`.

Use `list_variables("full", category = "debt")` to see the debt and
fund-balance entries, and `list_variables("full", category = "expenditure")`
for the capital outlay items.

## Important caveats

Read these before ranking or comparing districts on capital spending.

1. **Capital is not current spending.** Capital outlay is excluded from
   `exp_cur_total` by definition. Do not add it to current spending to get a
   "total" without knowing what you are combining.
2. **Capital is lumpy.** A district that builds a school records a large
   `exp_cap_total_pp` in one year and near-zero in the surrounding years.
   Single-year per-pupil capital rankings are misleading; use **3-5 year
   averages** instead.
3. **Bond-funded vs. pay-as-you-go.** High capital outlay may be financed by
   borrowing rather than current resources. Read `debt_lt_issued` and
   `fund_bal_bond` alongside the outlay flows to understand how construction was
   paid for.
4. **Interest, not principal.** `exp_debt_interest` is interest only. Principal
   retirement (`debt_lt_retired`) is repayment of a liability, not an
   expenditure.
5. **Interplay with the state-revenue adjustment.** edfinr nets state capital
   and debt aid out of `rev_state`; the pre-adjustment value is preserved in
   `rev_state_unadj` / `rev_state_unadj_pp`, and `c11_spike_flag` flags
   district-years where that adjustment spikes. Consider these when relating
   revenue to capital activity.
6. **`NA` is not zero.** A missing capital, debt, or fund-balance value means
   the item was not reported, not that the district spent or owed nothing.
7. **Stocks stay nominal.** Debt and fund-balance **stocks** (`debt_*`,
   `fund_bal_*`) are balance-sheet levels and are never CPI-adjusted, even when
   you pass `cpi_adj`. Capital outlay flows and `exp_debt_interest` *are*
   adjusted.

## Worked example: multi-year average capital per pupil

Because capital is lumpy, average it over several years. Here we adjust flows to
2023 dollars and keep districts with at least three years of data.

```{r multiyear-average}
oh <- get_finance_data(
  yr = "2019:2023", geo = "OH",
  dataset_type = "full", cpi_adj = "2023"
)

cap_multiyear <- oh |>
  group_by(ncesid, dist_name) |>
  summarize(
    n_years = sum(!is.na(exp_cap_total_pp)),
    avg_cap_pp = mean(exp_cap_total_pp, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(n_years >= 3) |>
  arrange(desc(avg_cap_pp))

head(cap_multiyear, 10)
```

Contrast that with a single-year snapshot to see how volatile one year can be:

```{r single-year-volatility}
oh |>
  filter(dist_name %in% head(cap_multiyear$dist_name, 5)) |>
  select(year, dist_name, exp_cap_total_pp) |>
  arrange(dist_name, year)
```

## Worked example: how was construction paid for?

Pair the outlay flow with debt issued and the bond fund balance. Debt and fund
balances are nominal, so this example leaves `cpi_adj` at its default.

```{r bond-vs-paygo}
oh_2023 <- get_finance_data(yr = "2023", geo = "OH", dataset_type = "full")

oh_2023 |>
  filter(!is.na(exp_cap_total)) |>
  arrange(desc(exp_cap_total)) |>
  transmute(
    dist_name,
    capital_outlay = exp_cap_total,
    lt_debt_issued = debt_lt_issued,
    bond_fund_balance = fund_bal_bond
  ) |>
  head(10)
```

Districts with large outlay and large `debt_lt_issued` are financing facilities
with borrowing; those with high outlay but little new debt are more likely using
current resources or accumulated bond-fund balances.

## Confirming stocks are not inflation-adjusted

`cpi_adj` scales the capital outlay flow but leaves the debt stock untouched:

```{r stocks-nominal}
raw <- get_finance_data(yr = "2019", geo = "OH", dataset_type = "full")
adj <- get_finance_data(yr = "2019", geo = "OH", dataset_type = "full", cpi_adj = "2023")

# capital outlay (a flow) is scaled up to 2023 dollars
head(adj$exp_cap_total / raw$exp_cap_total, 3)

# long-term debt outstanding (a stock) is identical in both
identical(adj$debt_lt_end, raw$debt_lt_end)
```

## See also

- The "CWIFT" vignette, if you want to compare labor costs across districts.
- The "CPI Adjustments" vignette for how flows are converted to constant dollars.
- `list_variables("full")` for the complete data dictionary.
