---
title: "Data Sources and Methodology"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Data Sources and Methodology}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE
)
```

```{r setup, include = FALSE}
library(edfinr)
library(dplyr)
```

## Overview

This vignette provides detailed information about the data sources and processing methods used to prepare the data used by the `edfinr` package. Understanding these details will help you interpret the data appropriately and inform analytical decisions.

Full data processing methods and scripts are available on GitHub via [bellwetherorg/edfinr_data_cleaning](https://github.com/bellwetherorg/edfinr_data_cleaning).

## Data Sources

This package provides access to education finance data from:

- [NCES CCD F-33 Data](https://nces.ed.gov/ccd/files.asp).
- NCES CCD Directory Data via the [Urban Institute's `educationdata` package](https://educationdata.urban.org/documentation/#r).
- [Census Bureau SAIPE Estimates](https://www.census.gov/programs-surveys/saipe.html)
- American Community Survey 5-Year Estimates via [`tidycensus` package](https://walker-data.com/tidycensus/).
- U.S. Bureau of Labor Statistics [Consumer Price Index for All Urban Consumers (CPI-U)](https://data.bls.gov/toppicks?survey=cu).
- NCES EDGE [Comparable Wage Index for Teachers (CWIFT)](https://nces.ed.gov/programs/edge/Economic/TeacherWage).

## Data Processing Methods

- Methodology based on process used by [`edbuildr`](https://github.com/EdBuild/edbuildr), which is detailed on a [methodology page](http://data.edbuild.org/) and in their [workshop documentation](http://viz.edbuild.org/workshops/edbuildr/).
- The [EdFund Data Dictionary](https://data-dictionary.ed-fund.org/) informs our handling of F-33 data.
- Revenue adjustments for payments to other school systems follows the approach used by Kristen Blagg, Emily Gutierrez, and Fanny Terrones in [Funding Flows: Which Students Receive a Greater Share of School Funding?](https://apps.urban.org/features/school-funding-trends/files/202204_K12_funding_technical_appendix.pdf)
- Inflation adjustments use an average of second half CPI-U of one year and first half CPI-U of the following year to align with the academic calendar.

## Data Processing Detail

### NCES F-33 Survey Data

Data source: NCES Common Core of Data text files of F-33 data from 2011-12 through 2022-23.

The crosswalk below maps every F-33-derived column in the full dataset to the
F-33 survey item(s) it is drawn from. It is generated from the package's data
dictionary, so it cannot drift from what `list_variables()` reports; single
codes are 1:1 mappings and formula strings describe simple combinations.

```{r f33-crosswalk}
list_variables("full") |>
  filter(source == "NCES F-33 Survey", !is.na(f33_item)) |>
  select(name, f33_item, description) |>
  knitr::kable()
```

In addition to the items above, the cleaning scripts read C11 (payments to
state governments), U11 (property sales), C24, L12, M12, and D11 to construct
the revenue adjustments described under "Revenue Adjustments" below. Variables
with an `NA` `f33_item` in `list_variables()` are either drawn from non-F-33
sources or are edfinr-adjusted measures whose construction is documented in
that section.

Adjustments applied during cleaning:

- Rename variables.
- Convert district names to title case.
- Ensure enrollment is a numeric variable.
- Replace `-1` and `-2` codes with `NA` values.

### CCD Directory Data

Data source: NCES CCD Directory data obtained via the
[educationdata](https://educationdata.urban.org/documentation/#r)
package.

Raw variables selected:

- Core district identifiers and location: state, ncesid, county, dist_name, state_leaid.
- Institutional details: lea_type, lea_type_id, urbanicity, congressional_dist.

Adjustments:

- Rename variables to more intuitive names.

### SAIPE Poverty Estimates

Data source: Census Bureau SAIPE Estimates.

Raw variables selected:

- Basic geographic and demographic fields: State Postal Code, State FIPS Code, District ID, Name
- Population estimates: Estimated Total Population, Estimated Population ages 5-17, and the estimated number of relevant children ages 5 to 17 living in poverty

Adjustments:

- Convert population fields to numeric
- Construct a combined NCES district identifier by concatenating state FIPS and District ID

### ACS 5-Year Estimates

Data source: American Community Survey 5-Year Estimates accessed via the
[`tidycensus`](https://walker-data.com/tidycensus/) package.

Raw variables selected:

- Economic indicators: Median household income (B19013_001), mean household income (aggregate household income divided by households), median property value (B25077_001), and the Gini index of income inequality (B19083_001).
- Household and labor characteristics: owner-occupied share of occupied housing (B25003), share of households receiving SNAP (B22003), and the unemployment rate among the civilian labor force (B23025).
- Educational attainment: Total population 25 years or older (B15003_001) and subsets of that population holding bachelor's degrees (B15003_022), master's degrees (B15003_023), professional degrees (B15003_024), and doctoral degrees (B15003_025).
- Data are pulled for different geographic breakdowns (unified, elementary, and secondary school districts).

Adjustments:

- Reshape data from long to wide format.
- Rename "GEOID" to a standard `ncesid` and ensure proper formatting of district identifiers.
- Convert estimates to numeric as needed.

### CPI

Data source: U.S. Bureau of Labor Statistics, specifically the Consumer Price Index for All Urban Consumers (CPI-U).

Raw variables selected:

- CPI time series data (specific variable names as provided in the raw file).

Adjustments:

- Calculate an averaged CPI value using the second half of one year and the first half of the following year to align with the academic calendar, with the 2011-12 school year as the baseline year.
- Clean and reformat CPI data for consistency across processing scripts.

### CWIFT (Comparable Wage Index for Teachers)

Data source: NCES EDGE [Comparable Wage Index for Teachers (CWIFT)](https://nces.ed.gov/programs/edge/Economic/TeacherWage), LEA-level releases.

Raw variables selected:

- District identifier (`LEAID`), the CWIFT estimate (`LEA_CWIFTEST`), and its standard error (`LEA_CWIFTSE`).

Adjustments and coverage:

- Each `CWIFT<yyyy>` release maps to edfinr fiscal year `yyyy`.
- FY2012-FY2014 have no CWIFT release and are returned as `NA`.
- FY2020 (no NCES release, owing to withheld ACS 2020 estimates) is interpolated as the mean of the FY2019 and FY2021 values for LEAs present in both years; the interpolated standard error is an approximation, not an NCES-published value.
- FY2023 is carried forward from FY2022 (no CWIFT2023 release as of the 2026-07-20 check).
- `cwift_imputed` flags interpolated or carried-forward values and `cwift_impute_method` records how each value was produced.

## Joining Data

- The joining process is implemented in the `08_edfinr_join_and_exclude.R` script (CWIFT is prepared in `07_cwift_clean.R` and joined there).
- Data from the F-33 survey, CCD Directory, ACS (unified, elementary, and secondary), SAIPE, and CWIFT sources are merged using left joins on shared district identifiers (ncesid) and fiscal year.
- The procedure ensures that each district record is enriched with revenue, expenditure, demographic, and economic data.

## Revenue Adjustments

Additional transformations are applied after the join:

- Capital expenditures and debt service (C11) are subtracted from state revenues. The unadjusted state revenue is preserved in `rev_state_unadj` / `rev_state_unadj_pp`, and `c11_spike_flag` marks district-years where this adjustment produces an anomalous spike.
- Property sales (U11) are subtracted from local revenues.
- For Texas local education agencies (LEAs) in school year 2012-13 and earlier, payments to state governments (L12) are subtracted from local revenues.
- Payments to other school systems (V91, V92, and Q11) are proportionally subtracted from local, state, and federal revenues.

## Exclusions

- Districts with enrollment less than 0 are removed.
- Districts with total revenue less than 0 are removed.
- Districts with an invalid LEA type (i.e. where lea_type_id is not one of 1, 2, 3, or 7) are excluded.
- Districts with invalid LEA/school level type (i.e. where schlev is not one of "01", "02", or "03", except for specified CA exceptions) are excluded.
- Districts where total revenue per-pupil is greater than $70,000 in school year 2011-12 dollars are excluded.
- Districts where total revenue per pupil is less than $500 in school year 2011-12 dollars are excluded.
- Connecticut LEAs consisting of semi-private high schools are removed (NCES IDs "0905371", "0905372", and "0905373").

## Changes from 0.1.x

Users comparing results against edfinr 0.1.x should be aware of these
methodology-relevant changes in 0.2.0:

- **`exp_cur_total` is sourced from TCURELSC.** In 0.1.x the total was the sum
  of the ESSA fund-type items (CE1 + CE2 + CE3), which several states skip
  entirely (all of Illinois and Minnesota through FY23; New York -- including
  New York City -- New Jersey, Massachusetts, Oregon, and others in earlier
  years) and which did not exist before FY16. Sourcing the total directly from
  the F-33 TCURELSC item makes `exp_cur_total`, `exp_cur_pp`, and
  `rev_exp_pp_diff` available for nearly all districts in every year 2012-2023.
  Where states did report the fund-type items, values shift slightly: the ESSA
  items exclude payments to private, charter, and other school systems, so the
  CE-sum differs from TCURELSC by more than 2% for roughly 40% of reporting
  districts. The fund-type split remains available as `exp_cur_st_loc`,
  `exp_cur_fed`, and `exp_cur_resa` (`NA` where unreported); those components
  should not be expected to sum exactly to `exp_cur_total`. Missing values
  still propagate to `NA` rather than being treated as zero.
- **`year` is returned as an integer** rather than a character column.
- **Hosted data format.** The hosted datasets are gzip-compressed Parquet
  files read with `nanoparquet` (previously `.rds`), and each year is hosted
  as a separate file so only the requested years are downloaded. This is
  transparent to callers.

## Data Notes and Cautions

Users should note the following when working with the `edfinr` datasets. For
worked examples of the diagnostic flags and comparability issues summarized
here, see the "Data Quality and Comparability" article on the package website.

- Some variables were originally coded with `-1` to indicate missing values; these have been replaced with `NA` during processing. An `NA` value means the item was not reported, not that it is zero.
- A wave of California charter schools became separate LEAs beginning in school year 2018-19, which sharply increases the number of California districts in the panel from 2019 onward. See the "Data Quality and Comparability" article for the full explanation and its implications for longitudinal analysis.
- The joined dataset represents a synthesis of data from multiple sources; discrepancies in source data formats may lead to minor variations.
- Inflation and adjustment factors (e.g., CPI adjustments) are based on averages and may not perfectly reflect local cost variations.
- Capital outlay is reported separately from current spending and is excluded from `exp_cur_total`. It is lumpy from year to year, so multi-year averages are recommended for cross-district comparison (see the "Capital and Facilities" article). Debt and fund-balance stocks (`debt_*`, `fund_bal_*`) are point-in-time balance-sheet levels and are never CPI-adjusted.
- CWIFT is a relative labor-cost index, not a price deflator, and has gaps that are imputed for some years (see the "CWIFT" article).
- **Caution is advised when comparing data across fiscal years due to potential differences in data collection and processing methods.**
