---
title: "Data Sources and Methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Sources and Methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This vignette provides detailed information about the data sources, processing methods, and quality controls used in the `edfinr` package. Understanding these details will help you interpret the data correctly and make appropriate analytical decisions.

```{r setup}
library(edfinr)
library(dplyr)
```


## Data Sources

This package provides access to education finance data from:

- [NCES CCD F-33 Data](https://nces.ed.gov/ccd/files.asp)
- NCES CCD Directory Data via the [Urban Institute's `educationdata` package](https://educationdata.urban.org/documentation/#r)
- [Census Bureau SAIPE Estimates](https://www.census.gov/programs-surveys/saipe.html)
- American Community Survey 5-Year Estimates via [`tidycensus` package](https://walker-data.com/tidycensus/)
- U.S Bureau of Labor Statistics [Consumer Price Index for All Urban Consumers (CPI-U)](https://data.bls.gov/toppicks?survey=cu)

## Data Processing Methods

- Methodology based on process used by [`edbuildr`](https://github.com/EdBuild/edbuildr), which is detailed on a [methodology page](http://data.edbuild.org/) and in their [workshop documentation](http://viz.edbuild.org/workshops/edbuildr/).
- The [EdFund Data Dictionary](https://data-dictionary.ed-fund.org/?_gl=1*199anoz*_ga*MTg3MDM3NDg2LjE3MzkzNzAzOTE.*_ga_TGH6XK399M*MTc0NDIyMzY3Ni43LjEuMTc0NDIyMzY4MC4wLjAuMA..) informs our handling of F-33 data.
- Revenue adjustments for payments to other school systems follows the approach used by Kristen Blagg, Emily Gutierrez, and Fanny Terrones in [Funding Flows: Which Students Receive a Greater Share of School Funding?](https://apps.urban.org/features/school-funding-trends/files/202204_K12_funding_technical_appendix.pdf).
- Inflation adjustments use an average of second half CPI-U of one year and first half CPI-U of the following year to align with the academic calendar.

## Data Processing Detail

### NCES F-33 Survey Data

Data source: NCES Common Core of Data text files of F-33 data from 2011-12 through 2021-22.

Raw variables selected:

- Basic information: state, leaid, name, yrdata, V33
- Revenue data: totalrev, tlocrev, tstrev, tfedrev
- Expenditure data: c11, u11, v91, v92, c24, l12, m12, d11, q11
- Current expenditure data: ce1, ce2, and ce3
- Detailed expenditure data: z32, z34, v93, v95, v02, k14, e13, z33, v10, e17, v11, v12, e07, v13, v14, e08, v15, v16, e09, v17, v18, v40, v21, v22, v45, v23, v24, v90, v37, v38, e11, v29, v30, v60, v32, v65, ae1, ae2, ae3, ae4, ae5, ae6, ae7, ae8

Adjustments:

- Rename variables
- Convert district names to title case
- Ensure enrollment is a numeric variable
- Replace `-1` and `-2` codes with `NA` values

### CCD Directory Data

Data source: NCES CCD Directory data obtained via the
[educationdata](https://educationdata.urban.org/documentation/#r)
package.

Raw variables selected:

- Core district identifiers and location: state, ncesid, county, dist_name, state_leaid
- Institutional details: lea_type, lea_type_id, urbanicity, congressional_dist

Adjustments:

- Rename variables to more intuitive names

### SAIPE Poverty Estimates

Data source: Census Bureau SAIPE Estimates

Raw variables selected:

- Basic geographic and demographic fields: State Postal Code, State FIPS Code, District ID, Name
- Population estimates: Estimated Total Population, Estimated Population 5-17, and the estimated number of relevant children 5 to 17 years old in poverty

Adjustments:

- Convert population fields to numeric
- Construct a combined NCES district identifier by concatenating state FIPS and District ID

### ACS 5-Year Estimates

Data source: American Community Survey 5-Year Estimates accessed via the
[`tidycensus`](https://walker-data.com/tidycensus/) package

Raw variables selected:

- Economic indicators: Median household income (B19013_001) and median property value (B25077_001)
- Educational attainment: Total population 25 years or older (B15003_001) and subsets of that population holding bachelor's degrees (B15003_022), master's degrees (B15003_023), professional degrees (B15003_024), and doctoral degrees (B15003_025).
- Data are pulled for different geographic breakdowns (unified, elementary, and secondary school districts)

Adjustments:

- Reshape data from long to wide format
- Rename “GEOID” to a standard `ncesid` and ensure proper formatting of district identifiers
- Convert estimates to numeric as needed

### CPI

Data source: U.S. Bureau of Labor Statistics, specifically the Consumer Price Index for All Urban Consumers (CPI-U)

Raw variables selected:

- CPI time series data (specific variable names as provided in the raw file)

Adjustments:

- Calculate an averaged CPI value using the second half of one year and the first half of the following year to align with the academic calendar, with the 2011-12 school year as the baseline year
- Clean and reformat CPI data for consistency across processing scripts

## Joining Data

- The joining process is implemented in the `07_edfinr_join_and_exclude.R` script.
- Data from the F-33 survey, CCD Directory, ACS (unified, elementary, and secondary), and SAIPE sources are merged using left joins on shared district identifiers (ncesid) and fiscal year.
- The procedure ensures that each district record is enriched with revenue, expenditure, demographic, and economic data.

## Revenue Adjustments

Additional transformations are applied after the join:
- Capital expenditures and debt service (C11) are subtrated from state revenues
- Property sales (U11) are subtracted from local revenues
- For Texas LEAs in 2012-13 and earlier, payments to state governments (L12) are subtracted from local revenues
- Payments to other school systems (V91, V92, and Q11) are proportionally subracted from local, state, and federal revenues

## Exclusions

- Districts with enrollment less than 0 are removed.
- Districts with total revenue less than 0 are removed.
- Districts with an invalid LEA type (i.e. where lea_type_id is not one of 1, 2, 3, or 7) are excluded.
- Districts with invalid LEA/school level type (i.e. where schlev is not one of "01", "02", or "03", except for specified CA exceptions) are excluded.
- Districts where total revenue per-pupil is greater than $70,000 in 2011-12 dollars are excluded.
- Districts where total revenue per pupil is less than $500 in 2011-12 dollars are excluded.
- Connecticut LEAs consisting of semi-private high schools are removed (NCES IDs "0905371", "0905372", and "0905373").
