---
title: "Community and Economic Context"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 8
)
```

## Introduction

The communities that fund and attend public schools are deeply connected to school finance. Alongside the F-33 finance items, `edfinr` joins each district to community measures from the American Community Survey (ACS) 5-Year Estimates, poverty estimates from the Census Bureau's Small Area Income and Poverty Estimates (SAIPE), and district land area from the Census Bureau's Gazetteer Files. This article explores those variables and shows how to use them for a typical analysis task: comparing districts' fiscal **capacity** (what a community can raise) with student **need**.

```{r setup, include = FALSE}
library(edfinr)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r, eval = FALSE}
library(edfinr)
library(dplyr)
library(tidyr)
library(ggplot2)
```

## The community variables

All of these are in the skinny dataset. The dictionary lists them with their
source table codes:

```{r variable-tour}
list_variables("skinny") |>
  filter(source %in% c("5-Year ACS Survey", "Census Bureau SAIPE", "Census Bureau Gazetteer")) |>
  select(name, source, description) |>
  knitr::kable()
```

Two timing caveats before using them:

1. **ACS values are 5-year averages.** The ACS estimates attached to a fiscal year summarize a five-year window, so they move slowly and lag sharp local changes (a plant closing, a housing boom). SAIPE poverty estimates are annual.
2. **Estimates carry sampling error.** ACS district-level estimates, especially for small districts, have nontrivial margins of error that `edfinr` does not carry. Treat small-district differences in `gini`, `snap_pct`, or `unemp_rate` as noisy.
3. **SAIPE counts resident children, not enrolled students.** `stpov_pct` measures poverty among children ages 5-17 who live within the district's boundaries. Where many resident children attend charter, private, or neighboring schools -- or where a charter LEA has no meaningful residential geography -- the resident and enrolled populations can differ substantially.

The examples below use the most recent year:

```{r download}
us_2023 <- get_finance_data(yr = "2023", geo = "all")
```

## Median vs. mean household income

`mhi` (median) and `mean_hhi` (mean) are different measures of community income. The mean is pulled upward by high-income households, so the mean-to-median ratio is a quick screen for skewed income distributions. This can be useful for identifying communities where a small wealthy population skews the mean income above the median income:

```{r income-skew}
us_2023 |>
  filter(!is.na(mhi), !is.na(mean_hhi), mhi > 0) |>
  mutate(mean_to_median = mean_hhi / mhi) |>
  filter(enroll >= 1000) |>
  arrange(desc(mean_to_median)) |>
  select(dist_name, state, enroll, mhi, mean_hhi, mean_to_median) |>
  head(10)
```

## Fiscal capacity: property wealth and local revenue

Local revenue depends heavily on the property tax base. Plotting per-pupil local revenue against median property value shows the relationship between the proxy for local fiscal capacity (MPV) and actual local revenue.

One caution before leaning on `mpv` as a capacity measure: it is the ACS median value of owner-occupied homes, a proxy for *residential* property wealth only. Formal fiscal-capacity measures use assessed (or equalized) valuation per pupil, which also counts commercial, industrial, and utility property and is scaled by enrollment. If your state publishes assessed valuation per pupil, use it; `mpv` is a nationally consistent fallback.

```{r capacity, fig.height = 5, fig.alt = "Scatterplot of median property value on a log scale versus local revenue per-pupil, colored by urbanicity, with smoothed trend lines showing local revenue rising with property wealth."}
us_2023 |>
  filter(!is.na(mpv), !is.na(rev_local_pp), mpv > 0, rev_local_pp > 0,
         !is.na(urbanicity)) |>
  ggplot(aes(x = mpv, y = rev_local_pp, size = enroll, color = urbanicity)) +
  geom_point(alpha = 0.2) +
  geom_smooth(se = FALSE) +
  scale_x_log10(labels = scales::label_dollar()) +
  scale_y_continuous(labels = scales::label_dollar()) +
  scale_size_area(max_size = 10) +
  labs(
    title = "Local Revenue Tracks Property Wealth, SY2022-23",
    subtitle = "Each point is a district; log 10 scale x-axis",
    x = "Median Property Value (Log 10 scale)",
    y = "Local Revenue Per-Pupil",
    color = "Urbanicity",
    size = "Enrollment"
  ) +
  theme_minimal()
```

## Student need: poverty and total revenue

The SAIPE-based `stpov_pct` measures the share of school-aged children in poverty. Whether total revenue rises with poverty -- that is, whether state and federal aid more than offset weaker local capacity -- varies sharply by state. Here is the national picture:

```{r need, fig.height = 5, fig.alt = "Scatterplot of student poverty rate versus total revenue per-pupil on a log scale for all districts in SY2022-23, with a nearly flat smoothed trend line."}
us_2023 |>
  filter(!is.na(stpov_pct), !is.na(rev_total_pp)) |>
  ggplot(aes(x = stpov_pct, y = rev_total_pp, size = enroll)) +
  geom_point(alpha = 0.15, size = 0.8) +
  geom_smooth(se = FALSE) +
  scale_x_continuous(labels = scales::label_percent()) +
  scale_y_continuous(labels = scales::label_dollar()) +
  scale_size_area(max_size = 10) +
  labs(
    title = "Student Poverty and Total Revenue Per-Pupil, SY2022-23",
    x = "Students in Poverty",
    y = "Total Revenue Per-Pupil",
    size = "Enrollment"
  ) +
  theme_minimal()
```

The national trend masks significant variation at the state level. It also compares nominal dollars: high-poverty urban districts sit disproportionately in high-wage labor markets, so a dollar buys less there than in a low-cost rural district. Cost-adjusted, within-state analyses can and do reach different conclusions than this raw national scatter. See the "CWIFT" article for the labor-cost adjustment. 

## Other indicators

`gini` (income inequality), `owner_pct` (owner-occupied housing), `snap_pct` (SNAP receipt), and `unemp_rate` (unemployment) round out the community picture. They are most useful as controls or descriptive context; their distributions differ enough across urbanicity to be worth checking before pooling:

```{r indicators, fig.height = 6, fig.alt = "Faceted boxplots of the Gini index, owner-occupancy rate, SNAP receipt rate, and unemployment rate by urbanicity, showing the distributions differ across city, suburban, town, and rural districts."}
us_2023 |>
  filter(!is.na(urbanicity)) |>
  select(urbanicity, gini, owner_pct, snap_pct, unemp_rate) |>
  pivot_longer(-urbanicity, names_to = "indicator", values_to = "value") |>
  filter(!is.na(value)) |>
  ggplot(aes(x = urbanicity, y = value)) +
  geom_boxplot(outlier.alpha = 0.1) +
  facet_wrap(~indicator, scales = "free_y") +
  labs(
    title = "Community Indicators by Urbanicity, SY2022-23",
    x = "Urbanicity", y = NULL
  ) +
  theme_minimal()
```

## Geographic sparsity: land area and student density

`urbanicity` classifies districts into four categories, but districts within the same category still span a wide range of physical sparsity. `s_per_sq_mi` (students per square mile, derived from `land_area_sq_mi`) offers a continuous counterpart: a rural district serving a compact town looks different from one spread across a sparsely populated county, even though both are labeled "Rural."

Three caveats before using it:

1. **`NA` by design, not missingness.** Districts without a Census boundary -- such as charter schools -- have no Gazetteer file match, so `land_area_sq_mi` and `s_per_sq_mi` are `NA` for them in every year.
2. **`NA`, never `Inf`.** Where `land_area_sq_mi` is zero or unavailable, `s_per_sq_mi` is `NA` rather than an infinite or undefined ratio.
3. **Vermont coverage gap.** Vermont's Act 46 district consolidation left many post-consolidation LEAs without a matching Gazetteer boundary; match rates there run roughly 7-12% for those years, versus 97%+ elsewhere. Restrict Vermont density analyses to FY2012-FY2015 or FY2022 onward.

Density varies by orders of magnitude across urbanicity categories, so a log scale can be useful:

```{r density, fig.height = 5, fig.alt = "Boxplots of students per square mile by urbanicity on a log scale, showing City districts are far denser than Suburb, Town, and Rural districts, each successively less dense."}
us_2023 |>
  filter(!is.na(urbanicity), !is.na(s_per_sq_mi)) |>
  ggplot(aes(x = urbanicity, y = s_per_sq_mi)) +
  geom_boxplot(outlier.alpha = 0.1) +
  scale_y_log10(labels = scales::label_comma()) +
  labs(
    title = "Student Density by Urbanicity, SY2022-23",
    x = "Urbanicity", y = "Students per Square Mile (Log 10 scale)"
  ) +
  theme_minimal()
```

## See also

- The "Data Quality and Comparability" article, including its notes on ACS timing and panel composition.
- The "Mapping School Finance Data" article to see these measures on a map.
