---
title: "Community and Economic Context"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 8
)
```

## Introduction

School finance rarely makes sense in isolation from the communities that fund
and attend the schools. Alongside the F-33 finance items, `edfinr` joins each
district to community measures from the American Community Survey (ACS) 5-Year
Estimates and poverty estimates from the Census Bureau's Small Area Income and
Poverty Estimates (SAIPE). This article explores those variables and shows how to
use them for the most common framing in the field: comparing districts' fiscal
**capacity** (what a community can raise) with student **need**.

```{r setup, include = FALSE}
library(edfinr)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r, eval = FALSE}
library(edfinr)
library(dplyr)
library(tidyr)
library(ggplot2)
```

## The community variables

All of these are in the skinny dataset. The dictionary lists them with their
source table codes:

```{r variable-tour}
list_variables("skinny") |>
  filter(source %in% c("5-Year ACS Survey", "Census Bureau SAIPE")) |>
  select(name, source, description) |>
  knitr::kable()
```

Two timing caveats before using them:

1. **ACS values are 5-year averages.** The ACS estimates attached to a fiscal
   year summarize a five-year window, so they move slowly and lag sharp local
   changes (a plant closing, a housing boom). SAIPE poverty estimates are
   annual.
2. **Estimates carry sampling error.** ACS district-level estimates,
   especially for small districts, have nontrivial margins of error that
   `edfinr` does not carry. Treat small-district differences in `gini`,
   `snap_pct`, or `unemp_rate` as noisy.
3. **SAIPE counts resident children, not enrolled students.** `stpov_pct`
   measures poverty among children ages 5-17 who live within the district's
   boundaries. Where many resident children attend charter, private, or
   neighboring schools -- or where a charter LEA has no meaningful
   residential geography -- the resident and enrolled populations can differ
   substantially.

The examples below use the most recent year:

```{r download}
us_2023 <- get_finance_data(yr = "2023", geo = "all")
```

## Median vs. mean household income

`mhi` (median) and `mean_hhi` (mean) answer different questions. The mean is
pulled upward by high-income households, so the mean-to-median ratio is a
quick screen for skewed income distributions -- useful for spotting
communities where a small wealthy population coexists with a much less
affluent majority:

```{r income-skew}
us_2023 |>
  filter(!is.na(mhi), !is.na(mean_hhi), mhi > 0) |>
  mutate(mean_to_median = mean_hhi / mhi) |>
  filter(enroll >= 1000) |>
  arrange(desc(mean_to_median)) |>
  select(dist_name, state, enroll, mhi, mean_hhi, mean_to_median) |>
  head(10)
```

## Fiscal capacity: property wealth and local revenue

Local revenue depends heavily on the property tax base. Plotting per-pupil
local revenue against median property value shows the capacity gradient 
and how it differs by urbanicity.

One caution before leaning on `mpv` as a capacity measure: it is the ACS
median value of owner-occupied homes, a proxy for *residential* property
wealth only. Formal fiscal-capacity measures use assessed (or equalized)
valuation per pupil, which also counts commercial, industrial, and utility
property and is scaled by enrollment. A district with modest homes and a
power plant or a shopping corridor is higher-capacity than `mpv` suggests.
Where your state publishes assessed valuation per pupil, prefer it; `mpv` is
a nationally consistent fallback.

```{r capacity, fig.height = 5, fig.alt = "Scatterplot of median property value versus local revenue per-pupil on log scales, colored by urbanicity, with smoothed trend lines showing local revenue rising with property wealth."}
us_2023 |>
  filter(!is.na(mpv), !is.na(rev_local_pp), mpv > 0, rev_local_pp > 0,
         !is.na(urbanicity)) |>
  ggplot(aes(x = mpv, y = rev_local_pp, color = urbanicity)) +
  geom_point(alpha = 0.2, size = 0.8) +
  geom_smooth(se = FALSE) +
  scale_x_log10(labels = scales::label_dollar()) +
  scale_y_log10(labels = scales::label_dollar()) +
  labs(
    title = "Local Revenue Tracks Property Wealth, SY2022-23",
    subtitle = "Each point is a district; log scales",
    x = "Median Property Value",
    y = "Local Revenue Per-Pupil",
    color = "Urbanicity"
  ) +
  theme_minimal()
```

## Student need: poverty and total revenue

The SAIPE-based `stpov_pct` measures the share of school-aged children in
poverty. Whether total revenue rises with poverty -- that is, whether state
and federal aid more than offset weaker local capacity -- varies sharply by
state. Here is the national picture:

```{r need, fig.height = 5, fig.alt = "Scatterplot of student poverty rate versus total revenue per-pupil on a log scale for all districts in SY2022-23, with a nearly flat smoothed trend line."}
us_2023 |>
  filter(!is.na(stpov_pct), !is.na(rev_total_pp)) |>
  ggplot(aes(x = stpov_pct, y = rev_total_pp)) +
  geom_point(alpha = 0.15, size = 0.8) +
  geom_smooth(se = FALSE) +
  scale_x_continuous(labels = scales::label_percent()) +
  scale_y_log10(labels = scales::label_dollar()) +
  labs(
    title = "Student Poverty and Total Revenue Per-Pupil, SY2022-23",
    x = "Students in Poverty",
    y = "Total Revenue Per-Pupil (log scale)"
  ) +
  theme_minimal()
```

Resist reading the flat national line as a statement about school funding
progressivity. It pools fifty different funding systems, and it compares
nominal dollars: high-poverty urban districts sit disproportionately in
high-wage labor markets, so a dollar buys less there than in a low-cost rural
district. Cost-adjusted, within-state analyses can and do reach different conclusions
than this raw national scatter. See the "CWIFT" article for the labor-cost
adjustment, and prefer the within-state framing below for any claim about how
a funding system treats high-poverty districts.

## Need vs. capacity within a state

Cross-state comparisons mix different funding systems, so the need-capacity
framing is sharpest within one state. Classifying districts against their
state's median poverty and median property value yields four quadrants; the
high-need, low-capacity quadrant is where funding systems are tested:

```{r quadrants, fig.height = 5.5, fig.alt = "Scatterplot of Ohio districts' median property value versus student poverty rate, colored by total revenue per-pupil, with dashed lines at the state medians dividing districts into four need-capacity quadrants."}
oh_2023 <- us_2023 |>
  filter(state == "OH", !is.na(stpov_pct), !is.na(mpv))

oh_quad <- oh_2023 |>
  mutate(
    need = ifelse(stpov_pct > median(stpov_pct), "Higher need", "Lower need"),
    capacity = ifelse(mpv > median(mpv), "Higher capacity", "Lower capacity"),
    quadrant = paste(need, capacity, sep = ", ")
  )

# how does total revenue differ across the quadrants?
oh_quad |>
  group_by(quadrant) |>
  summarize(
    n_districts = n(),
    med_rev_total_pp = median(rev_total_pp, na.rm = TRUE),
    med_rev_local_pp = median(rev_local_pp, na.rm = TRUE),
    med_rev_state_pp = median(rev_state_pp, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(oh_quad, aes(x = mpv, y = stpov_pct, color = rev_total_pp)) +
  geom_point(alpha = 0.7) +
  geom_vline(xintercept = median(oh_2023$mpv), linetype = "dashed") +
  geom_hline(yintercept = median(oh_2023$stpov_pct), linetype = "dashed") +
  scale_x_log10(labels = scales::label_dollar()) +
  scale_y_continuous(labels = scales::label_percent()) +
  scale_color_viridis_c(labels = scales::label_dollar()) +
  labs(
    title = "Need vs. Capacity in Ohio Districts, SY2022-23",
    subtitle = "Dashed lines mark state medians",
    x = "Median Home Value (capacity proxy)",
    y = "Students in Poverty (need)",
    color = "Total Revenue\nPer-Pupil"
  ) +
  theme_minimal()
```

## The other indicators

`gini` (income inequality), `owner_pct` (owner-occupied housing),
`snap_pct` (SNAP receipt), and `unemp_rate` (unemployment) round out the
community picture. They are most useful as controls or descriptive context;
their distributions differ enough across urbanicity to be worth checking
before pooling:

```{r indicators, fig.height = 6, fig.alt = "Faceted boxplots of the Gini index, owner-occupancy rate, SNAP receipt rate, and unemployment rate by urbanicity, showing the distributions differ across city, suburban, town, and rural districts."}
us_2023 |>
  filter(!is.na(urbanicity)) |>
  select(urbanicity, gini, owner_pct, snap_pct, unemp_rate) |>
  pivot_longer(-urbanicity, names_to = "indicator", values_to = "value") |>
  filter(!is.na(value)) |>
  ggplot(aes(x = urbanicity, y = value)) +
  geom_boxplot(outlier.alpha = 0.1) +
  facet_wrap(~indicator, scales = "free_y") +
  labs(
    title = "Community Indicators by Urbanicity, SY2022-23",
    x = "Urbanicity", y = NULL
  ) +
  theme_minimal()
```

## See also

- The "Data Quality and Comparability" article, including its notes on ACS timing and panel composition.
- The "Mapping School Finance Data" article to see these measures on a map.
- The "Basic usage of edfinr" vignette for `list_variables()` and the data dictionary.
