---
title: "COVID Relief Spending"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 8
)
```

## Introduction

Beginning in FY2020, the F-33 survey added items tracking district expenditures
of temporary federal COVID-19 assistance funds (ESSER and related programs).
`edfinr`'s **full** dataset carries these as the `exp_covid_*` variables, which
let you see how much pandemic aid districts spent, on what, and how that
spending wound down as the programs expired.

```{r setup, include = FALSE}
library(edfinr)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r, eval = FALSE}
library(edfinr)
library(dplyr)
library(tidyr)
library(ggplot2)
```

The eight variables map to F-33 items AE1-AE8
(see `list_variables("full")`):

- `exp_covid_total` (AE1) -- total expenditures from COVID-19 federal
  assistance funds.
- `exp_covid_instr` (AE2) -- instructional expenditures.
- `exp_covid_supp` (AE3) -- support services expenditures.
- `exp_covid_cap_out` (AE4) -- capital outlay expenditures.
- `exp_covid_tech_supp` (AE5) -- technology-related supplies and purchased
  services.
- `exp_covid_tech_equip` (AE6) -- technology-related equipment.
- `exp_covid_supp_plant` (AE7) -- operation and maintenance of plant
  (from FY2021).
- `exp_covid_food` (AE8) -- food services operations (from FY2021).

## Important caveats

1. **Full dataset only.** The `exp_covid_*` variables are not in the skinny
   dataset; request `dataset_type = "full"`.
2. **Availability.** AE1-AE6 begin in FY2020; AE7 and AE8 begin in FY2021
   (`first_yr_avail` in the dictionary records this).
3. **A lens, not an addition.** The AE items report spending *from* federal
   COVID assistance funds. That spending also appears in the regular
   expenditure items, so do not add `exp_covid_*` to the standard expenditure
   variables -- that double counts.
4. **`NA` is not zero.** Districts that did not report an item show `NA`.
   Reporting completeness for these items varies across states, especially in
   FY2020, so treat state-level comparisons with care.
5. **CPI adjustment applies.** Like other expenditure flows, the `exp_covid_*`
   variables are scaled when you pass `cpi_adj`. The examples below stay
   nominal to match how federal awards are usually described.

## The arc of relief spending

```{r download}
covid <- get_finance_data(yr = "2020:2023", geo = "all", dataset_type = "full")

national <- covid |>
  filter(!is.na(exp_covid_total), !is.na(exp_cur_total)) |>
  group_by(year) |>
  summarize(
    covid_total = sum(exp_covid_total),
    cur_total = sum(exp_cur_total),
    enroll = sum(enroll, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    covid_pp = covid_total / enroll,
    covid_share = covid_total / cur_total
  )

national |>
  mutate(
    covid_total_b = round(covid_total / 1e9, 1),
    covid_pp = round(covid_pp),
    covid_share = round(100 * covid_share, 1)
  ) |>
  select(year, covid_total_b, covid_pp, covid_share)
```

The table shows total reported COVID-fund spending (in billions), per-pupil
spending, and COVID-fund spending as a share of current expenditure among
districts reporting both items.

## What did districts spend relief funds on?

The component items break spending into instruction, support services, capital
outlay, technology, plant operations, and food service. Components do not
mechanically sum to `exp_covid_total` -- districts report each item separately
and some spending falls outside the listed categories -- so shares below are
of the summed components, by year.

```{r composition, fig.height = 5}
component_labels <- c(
  exp_covid_instr = "Instruction",
  exp_covid_supp = "Support services",
  exp_covid_cap_out = "Capital outlay",
  exp_covid_tech_supp = "Technology supplies/services",
  exp_covid_tech_equip = "Technology equipment",
  exp_covid_supp_plant = "Plant operations",
  exp_covid_food = "Food services"
)

covid |>
  select(year, all_of(names(component_labels))) |>
  pivot_longer(-year, names_to = "component", values_to = "amount") |>
  filter(!is.na(amount)) |>
  group_by(year, component) |>
  summarize(total = sum(amount), .groups = "drop_last") |>
  mutate(share = total / sum(total)) |>
  ungroup() |>
  mutate(component = component_labels[component]) |>
  ggplot(aes(x = year, y = share, fill = component)) +
  geom_col() +
  scale_y_continuous(labels = scales::label_percent()) +
  labs(
    title = "Composition of Reported COVID-Fund Spending by Year",
    subtitle = "Shares of summed component items (AE2-AE8), nominal dollars",
    x = "Year", y = "Share of Component Spending", fill = "Component"
  ) +
  theme_minimal()
```

## How did relief spending vary across districts?

Per-pupil relief spending in FY2023 varied widely across districts, reflecting
both the poverty-weighted allocation formulas and differences in spend-down
pacing:

```{r distribution, fig.height = 4.5}
covid |>
  filter(year == 2023, !is.na(exp_covid_total), enroll >= 200,
         !is.na(urbanicity)) |>
  mutate(covid_pp = exp_covid_total / enroll) |>
  ggplot(aes(x = urbanicity, y = covid_pp)) +
  geom_boxplot(outlier.alpha = 0.15) +
  scale_y_continuous(labels = scales::label_dollar()) +
  coord_cartesian(ylim = c(0, 5000)) +
  labs(
    title = "Per-Pupil COVID-Fund Spending by Urbanicity, FY2023",
    subtitle = "Districts with 200+ students; y-axis truncated at $5,000",
    x = "Urbanicity", y = "COVID-Fund Spending Per-Pupil"
  ) +
  theme_minimal()
```

## Tracking the wind-down in large districts

For any given district, the FY2020-FY2023 series shows the ramp-up and
spend-down of relief funds:

```{r wind-down, fig.height = 5}
largest_ids <- covid |>
  filter(year == 2023) |>
  slice_max(enroll, n = 6) |>
  pull(ncesid)

covid |>
  filter(ncesid %in% largest_ids, !is.na(exp_covid_total)) |>
  mutate(covid_pp = exp_covid_total / enroll) |>
  ggplot(aes(x = year, y = covid_pp)) +
  geom_line() +
  geom_point(size = 1) +
  facet_wrap(~dist_name, ncol = 2) +
  scale_y_continuous(labels = scales::label_dollar()) +
  labs(
    title = "Per-Pupil COVID-Fund Spending in the Six Largest Districts",
    subtitle = "FY2020-FY2023, nominal dollars",
    x = "Year", y = "COVID-Fund Spending Per-Pupil"
  ) +
  theme_minimal()
```

## See also

- The "Data Quality and Comparability" article for missingness and reporting caveats that apply doubly to these items.
- The "CPI Adjustments" vignette if you need relief spending in constant dollars.
- `list_variables("full", category = "expenditure")` for the full expenditure dictionary.
