---
title: "CWIFT"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 8
)
```

## What CWIFT is

The **Comparable Wage Index for Teachers (CWIFT)** is a National Center for
Education Statistics (NCES) EDGE index of how much it costs to employ
college-educated workers in a given labor market, relative to the national
average. It lets you account for the fact that a dollar buys less teacher labor
in a high-wage metro than in a low-cost rural area.

```{r setup, include = FALSE}
library(edfinr)
library(dplyr)
library(ggplot2)
```

```{r, eval = FALSE}
library(edfinr)
library(dplyr)
library(ggplot2)
```

`edfinr` provides four CWIFT columns:

- `cwift_est` -- the index value (roughly centered on 1.0; above 1.0 means
  higher-than-average labor costs). In the skinny dataset.
- `cwift_imputed` -- `TRUE` if the value was imputed rather than observed. In
  the skinny dataset.
- `cwift_se` -- the standard error of the estimate (full dataset).
- `cwift_impute_method` -- how the value was produced: `"observed"`,
  `"interpolated_2019_2021"`, or `"carried_forward_2022"` (full dataset).

## When *not* to use CWIFT

CWIFT is a **labor-cost index, not a general price deflator.** Use `cpi_adj`
(see the "CPI Adjustments" vignette) to convert dollars across *time*; use CWIFT
to compare labor costs across *places* in the same year. It does not deflate
construction costs or non-labor inputs, and combining it with CPI is a two-step
adjustment, not a single one (see below).

## Coverage and imputation

CWIFT is not available for every edfinr year. FY2012-FY2014 predate the series,
FY2020 was not published (the ACS 2020 estimates were withheld), and no FY2023
release existed at the time of processing.

```{r coverage}
cov <- get_finance_data(yr = "all", dataset_type = "skinny") |>
  group_by(year) |>
  summarize(
    n = n(),
    observed = sum(cwift_imputed == FALSE, na.rm = TRUE),
    imputed = sum(cwift_imputed == TRUE, na.rm = TRUE),
    missing = sum(is.na(cwift_est)),
    .groups = "drop"
  )

cov
```

The imputation rules:

- **FY2012-FY2014:** no release; all four columns are `NA`.
- **FY2020:** interpolated as the mean of FY2019 and FY2021 for LEAs present in
  both years (`cwift_impute_method == "interpolated_2019_2021"`). The
  interpolated `cwift_se` is an approximation, not an NCES-published value.
- **FY2023:** carried forward from FY2022
  (`cwift_impute_method == "carried_forward_2022"`).

Treat imputed values with more caution than observed ones; `cwift_imputed`
makes it easy to filter them out.

## How is CWIFT distributed?

Before ranking districts, it helps to see the index's spread. Labor costs rise
with urbanization, so the distribution shifts upward from rural districts to
cities. This example uses FY2022, the most recent observed (non-imputed) year.

```{r distribution, fig.height = 4.5}
us_2022 <- get_finance_data(yr = "2022", geo = "all", dataset_type = "full")

us_2022 |>
  filter(!is.na(cwift_est), !is.na(urbanicity)) |>
  ggplot(aes(x = urbanicity, y = cwift_est)) +
  geom_boxplot(outlier.alpha = 0.2) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  labs(
    title = "CWIFT by Urbanicity, FY2022",
    subtitle = "Dashed line marks the national average (1.0)",
    x = "Urbanicity",
    y = "CWIFT Estimate"
  ) +
  theme_minimal()
```

## High-cost and low-cost districts

```{r high-low}
d <- get_finance_data(yr = "2023", geo = "all", dataset_type = "full")

# highest-cost labor markets among larger districts
d |>
  filter(!is.na(cwift_est), enroll > 10000) |>
  arrange(desc(cwift_est)) |>
  select(dist_name, state, cwift_est, cwift_se, cwift_impute_method) |>
  head(5)

# lowest-cost labor markets among larger districts
d |>
  filter(!is.na(cwift_est), enroll > 10000) |>
  arrange(cwift_est) |>
  select(dist_name, state, cwift_est, cwift_se, cwift_impute_method) |>
  head(5)
```

## Reading `cwift_se`

`cwift_se` is the standard error of `cwift_est`; smaller values mean a more
precise estimate. It is most useful when comparing two districts whose point
estimates are close -- if their intervals overlap heavily, treat them as
similar. Remember that the standard error for interpolated FY2020 values is an
approximation.

Plotting estimates with intervals of plus or minus 1.96 standard errors makes
the overlap visible. Among Ohio's largest districts, several point estimates
that differ in the second decimal place have intervals that overlap
substantially, so the data cannot distinguish their labor costs:

```{r se-intervals, fig.height = 5}
us_2022 |>
  filter(state == "OH", !is.na(cwift_est), !is.na(cwift_se)) |>
  slice_max(enroll, n = 15) |>
  ggplot(aes(x = cwift_est, y = reorder(dist_name, cwift_est))) +
  geom_errorbar(
    aes(xmin = cwift_est - 1.96 * cwift_se, xmax = cwift_est + 1.96 * cwift_se),
    width = 0.3
  ) +
  geom_point() +
  labs(
    title = "CWIFT Estimates with 95% Intervals, 15 Largest Ohio Districts, FY2022",
    x = "CWIFT Estimate (±1.96 SE)",
    y = NULL
  ) +
  theme_minimal()
```

## Recipe: labor-cost-adjusted per-pupil dollars

`edfinr` ships the raw index rather than precomputed adjusted columns, so you
can divide any per-pupil dollar figure by `cwift_est` to express it in
labor-cost-adjusted terms. Districts in expensive labor markets look relatively
lower after adjustment.

```{r recipe}
d |>
  filter(!is.na(cwift_est), cwift_est > 0) |>
  mutate(rev_total_pp_cwift = rev_total_pp / cwift_est) |>
  arrange(desc(cwift_est)) |>
  select(dist_name, state, rev_total_pp, cwift_est, rev_total_pp_cwift) |>
  head(5)
```

## Combining CWIFT with CPI without double-counting

The two adjustments answer different questions, so apply them in sequence:

1. Use `cpi_adj` to put dollars from different years into constant dollars
   (adjust across time).
2. Divide the resulting per-pupil dollars by `cwift_est` to adjust for labor
   costs (adjust across place).

```{r combine}
la_ny <- get_finance_data(
  yr = "2022", geo = "CA,NY",
  dataset_type = "full", cpi_adj = "2023"
) |>
  filter(!is.na(cwift_est), cwift_est > 0, enroll > 50000) |>
  mutate(
    rev_total_pp_2023 = rev_total_pp,                     # already in 2023 dollars
    rev_total_pp_2023_cwift = rev_total_pp / cwift_est    # then adjust for labor cost
  ) |>
  select(dist_name, state, rev_total_pp_2023, cwift_est, rev_total_pp_2023_cwift)

head(la_ny, 10)
```

Do not CPI-adjust `cwift_est` itself -- it is an index, not a dollar amount, and
`get_finance_data()` never scales it.

## See also

- The "CPI Adjustments" vignette for time-based inflation adjustment.
- The "Capital and Facilities" article for capital, debt, and fund balances.
- The "Mapping School Finance Data" article to put `cwift_est` on a map.
- `list_variables("full", category = "cwift")` for the CWIFT dictionary entries.
