---
title: "Data Quality and Comparability"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 8
)
```

## Introduction

School finance data are messy in ways that summary statistics hide: districts
might skip survey items, states change how they report aid, and the set of districts
in the panel shifts over time. `edfinr` ships several diagnostic variables that
make these issues visible, and this article shows how to use them. If you plan
to rank districts, build a time series, or compare states, read this first.

```{r setup, include = FALSE}
library(edfinr)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r, eval = FALSE}
library(edfinr)
library(dplyr)
library(tidyr)
library(ggplot2)
```

The examples below use the full national panel:

```{r download}
us <- get_finance_data(yr = "all", geo = "all", dataset_type = "skinny")
```

## `NA` is not zero

Some F-33 items were originally coded `-1` or `-2` to indicate missing values;
`edfinr` recodes these to `NA` during processing. An `NA` therefore means the
item was **not reported**, not that the district spent or received nothing.
Dropping `NA`s silently (for example, with `na.rm = TRUE` in a ranking) keeps
incompletely-reporting districts in your analysis with partial values, which
can be worse than excluding them; decide explicitly.

Since 0.2.0, `edfinr` also detects a second F-33 missingness convention:
some items are **zero-filled** when missing, with a companion data-item flag
(`FL_* = "M"`) marking them unreported. The cleaning step converts these
flagged zero-fills to `NA` for the COVID (`exp_covid_*`), capital-detail,
debt, fund-balance, CE fund-type, and expenditure-detail items. The largest
effect is on the COVID columns: all of New York (including NYC) in every year
and roughly a third to half of California districts from FY2021 onward never
reported them, and those districts previously appeared as spending exactly
$0. Genuine reported zeros are preserved. Missingness
also **propagates** through derived measures rather than being treated as
zero. The share of district-years affected varies by column:

```{r na-by-year}
us |>
  group_by(year) |>
  summarize(
    n = n(),
    pct_na_exp_cur = round(100 * mean(is.na(exp_cur_total)), 1),
    pct_na_cap = round(100 * mean(is.na(exp_cap_total)), 1),
    .groups = "drop"
  )
```

`exp_cur_total` is near-fully populated in every year: it is sourced from the
F-33 summary item TCURELSC, not from the ESSA fund-type items (CE1/CE2/CE3),
which whole states skip and which only begin in FY2016 (CE3, `exp_cur_resa`,
in FY2018). The fund-type components (`exp_cur_st_loc`, `exp_cur_fed`, `exp_cur_resa`)
do carry those coverage boundaries; the dictionary's `first_yr_avail` column 
records them (see "Variable availability by year" below). Summary items like
TCURELSC carry no data-item flags, so the flag-aware cleaning above cannot
detect zero-filled missingness there -- another reason to treat exact zeros
in totals with mild suspicion.

## The state-revenue adjustment and `c11_spike_flag`

Following the EdBuild methodology, `edfinr` nets capital and debt-related state
aid (F-33 item C11) out of `rev_state`, so that `rev_state` approximates
operating support. The pre-adjustment values are preserved in
`rev_state_unadj` and `rev_state_unadj_pp`, and the netted-out C11 amount
itself ships as `rev_state_cap_debt` (in both datasets), so the adjustment can
be reconstructed directly: `rev_state_unadj - rev_state_cap_debt - rev_state`
equals the state share of the other-system-payment adjustment. Because
`rev_state_cap_debt` feeds the adjustment arithmetic, it is zero-filled, not
`NA`, where districts did not report.

The identity is easiest to see in districts where the adjustment is large.
Here are the three Ohio districts with the most state capital and debt aid in
FY2023; the residual after subtracting `rev_state_cap_debt` and `rev_state`
from the unadjusted total is the state share of the other-system-payment
adjustment:

```{r reconstruction}
us |>
  filter(year == 2023, state == "OH") |>
  slice_max(rev_state_cap_debt, n = 3) |>
  select(dist_name, rev_state_unadj, rev_state_cap_debt, rev_state) |>
  mutate(state_share_osp = rev_state_unadj - rev_state_cap_debt - rev_state)
```

In a small number of district-years, the C11 adjustment produces an anomalous
spike -- typically when a large one-time capital grant flows through -- and
`c11_spike_flag` marks them:

```{r spike-counts}
us |>
  group_by(year) |>
  summarize(flagged = sum(c11_spike_flag, na.rm = TRUE), .groups = "drop")

# which states account for the most flagged district-years?
us |>
  filter(c11_spike_flag) |>
  count(state, sort = TRUE) |>
  head(10)
```

For a flagged district, the adjusted state-revenue series can dip or jump in
ways the unadjusted series does not. Comparing the two makes the artifact
obvious. Here we pick the largest district with multiple flagged years:

```{r spike-example, fig.height = 4.5, fig.alt = "Line chart comparing adjusted and unadjusted state revenue per-pupil for one district from 2012 to 2023, with triangle markers on the years flagged by c11_spike_flag where the two series diverge sharply."}
example_id <- us |>
  filter(c11_spike_flag) |>
  count(ncesid, wt = enroll, sort = TRUE) |>
  slice(1) |>
  pull(ncesid)

us |>
  filter(ncesid == example_id) |>
  select(year, dist_name, c11_spike_flag, rev_state_pp, rev_state_unadj_pp) |>
  pivot_longer(
    cols = c(rev_state_pp, rev_state_unadj_pp),
    names_to = "series", values_to = "value"
  ) |>
  mutate(series = ifelse(
    series == "rev_state_pp", "Adjusted (rev_state_pp)", "Unadjusted (rev_state_unadj_pp)"
  )) |>
  ggplot(aes(x = year, y = value, color = series)) +
  geom_line() +
  geom_point(aes(shape = c11_spike_flag), size = 2.5) +
  scale_x_continuous(breaks = seq(2013, 2023, 2)) +
  scale_y_continuous(labels = scales::label_dollar()) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17)) +
  labs(
    title = "Adjusted vs. Unadjusted State Revenue Per-Pupil",
    subtitle = "Triangles mark years where c11_spike_flag is TRUE",
    x = "Year", y = "State Revenue Per-Pupil",
    color = NULL, shape = "Flagged"
  ) +
  theme_minimal()
```

Practical guidance: when analyzing state revenue over time, check
`c11_spike_flag` for your districts of interest, inspect `rev_state_cap_debt`
to see how large the adjustment actually is, and consider using
`rev_state_unadj_pp` (or excluding flagged years) where the adjustment
dominates the series.

## Pass-through districts and `osp_pct`

Some districts pass a large share of the revenue they receive on to other
school systems (charter schools, other LEAs, private placements). `edfinr`
proportionally subtracts these payments from revenues, and `osp_pct` reports
the share of unadjusted total revenue paid to other systems, so you can see
how much the adjustment matters for any district:

```{r osp}
us_2023 <- us |> filter(year == 2023)

# how common are large pass-through shares?
us_2023 |>
  summarize(
    n = sum(!is.na(osp_pct)),
    over_10_pct = sum(osp_pct > 0.10, na.rm = TRUE),
    over_25_pct = sum(osp_pct > 0.25, na.rm = TRUE)
  )

# the largest pass-through districts, among districts with 1,000+ students
us_2023 |>
  filter(enroll >= 1000) |>
  arrange(desc(osp_pct)) |>
  select(dist_name, state, enroll, osp_pct, rev_total_pp) |>
  head(10)
```

For high-`osp_pct` districts, per-pupil figures describe the students the
district serves directly, after the pass-through dollars are removed. When your
question concerns all dollars a district handles, use the `rev_*_unadj`
variables instead.

One caution: unlike the flag-cleaned expenditure items above, `osp_pct` and
the underlying `exp_pay_*` columns deliberately retain zero-filled values
where states did not report, because they feed the revenue-adjustment
arithmetic. An `osp_pct` of exactly zero can therefore mean either no
pass-through or no report.

## Panel composition: districts enter and exit

The set of districts is not constant across years, and treating the panel as
balanced will bias trend analyses. The largest single discontinuity is
California: beginning in school year 2018-19, a wave of California charter
schools switched to independent reporting and were assigned
their own NCES LEA IDs for the first time. Once in the NCES LEA universe,
those new charter-LEAs automatically show up in the F-33 finance survey and in
the Census SAIPE and ACS school-district products, which mirror NCES LEA
boundaries. The result is a jump in California district counts from 2019
onward that reflects reporting structure, not new schools:

```{r panel-plot, fig.height = 4.5, fig.alt = "Two-panel line chart of district counts by year for California and all other states; the California count jumps sharply in 2019 when charter schools became separately reported LEAs."}
us |>
  mutate(group = ifelse(state == "CA", "California", "All other states")) |>
  count(group, year) |>
  ggplot(aes(x = year, y = n)) +
  geom_line() +
  geom_point(size = 1) +
  facet_wrap(~group, scales = "free_y") +
  scale_x_continuous(breaks = seq(2013, 2023, 2)) +
  labs(
    title = "Districts in the edfinr Panel by Year",
    subtitle = "The 2019 jump in California reflects charter schools becoming separate LEAs",
    x = "Year", y = "Districts"
  ) +
  theme_minimal()
```

For longitudinal work, consider a **consistent sample**: keep only districts
present in every year of your window.

```{r consistent-sample}
window <- us |> filter(year >= 2019)

consistent_ids <- window |>
  count(ncesid) |>
  filter(n == n_distinct(window$year)) |>
  pull(ncesid)

length(consistent_ids)
```

Two 0.2.0 changes also altered the panel's composition relative to earlier
releases. First, directory attributes (name, county, LEA type, urbanicity)
now come from the same school year as the fiscal year they describe rather
than the following year's vintage; among other effects, this restored the
final operating year of districts that closed, which earlier releases
silently dropped. Second, Massachusetts regional school districts, which CCD
miscoded as service agencies for the SY2011-12 through SY2015-16 directory
vintages, are restored for FY2012-FY2015 via a vetted list of 60 districts
covering roughly 107,000-110,000 students per year. One trap comes with the
restoration: those rows (and the FY2016 rows) carry `lea_type_id` 4
("Service agency") because the package reports what the source vintage said,
so filtering Massachusetts years 2012-2016 on `lea_type_id` will drop real
regional districts. Row counts rise slightly in every year relative to
0.1.x.

Also remember that `edfinr` excludes some districts by construction (invalid
LEA types, extreme per-pupil revenue, zero or negative enrollment or
revenue); the "Data Sources and Methodology" vignette documents the
exclusion rules.

## Variable availability by year

Not every variable spans the full 2012-2023 panel. The dictionary's
`first_yr_avail` column records when each variable enters:

```{r availability}
list_variables("full") |>
  filter(first_yr_avail != "2012") |>
  select(name, first_yr_avail, description) |>
  arrange(first_yr_avail, name) |>
  knitr::kable()
```

A multi-year average or trend that crosses one of these boundaries will mix
real change with a coverage change. CWIFT adds a further wrinkle: FY2020 and
FY2023 values are imputed rather than observed (see the "CWIFT" article).

## Urbanicity: condensed vs. raw locale codes

The `urbanicity` factor condenses the NCES urban-centric locale codes into
four categories (City, Suburb, Town, Rural). The raw 12-category codes are
also available as `urbanicity_raw` (numeric) and `urbanicity_raw_cat`
(labeled factor):

```{r urbanicity}
us_2023 |>
  count(urbanicity_raw_cat, urbanicity) |>
  arrange(urbanicity, urbanicity_raw_cat)
```

The condensed categories are right for most summaries, but the raw codes
matter when the distinction *within* a category is the point -- for example,
"Rural: Fringe" districts sit at metro edges and often look more like suburbs
than like "Rural: Remote" districts.

## Comparability checklist

Before publishing numbers built on `edfinr` data:

1. **Check missingness** for your variables and years; decide explicitly how
   to treat `NA` district-years rather than relying on `na.rm = TRUE`.
2. **Check `first_yr_avail`** for every variable in a multi-year analysis.
3. **Check `c11_spike_flag`** (and compare against `rev_state_unadj_pp`) when
   state revenue is central to the analysis.
4. **Check `osp_pct`** when high-pass-through districts could distort
   per-pupil comparisons.
5. **Use a consistent sample** (or justify not doing so) for trend analyses,
   especially any that include California.
6. **Use CPI adjustment** for any multi-year dollar comparison (see the "CPI
   Adjustments" vignette), and remember debt and fund-balance stocks stay
   nominal.

## See also

- The "Data Sources and Methodology" vignette for the processing pipeline, exclusion rules, and the F-33 crosswalk.
- The "CWIFT" article for imputation flags on the wage index.
- The "Capital and Facilities" article for lumpiness in capital outlay.
