---
title: "Data Quality and Comparability"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 8
)
```

## Introduction

School finance data are messy in ways that summary statistics hide: districts and states might change how they report revenues and expenditures. Meanwhile, the set of districts in the panel shifts over time. `edfinr` ships several diagnostic variables that make these issues visible, and this article shows how to use them.

```{r setup, include = FALSE}
library(edfinr)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r, eval = FALSE}
library(edfinr)
library(dplyr)
library(tidyr)
library(ggplot2)
```

The examples below use the full national panel:

```{r download}
us <- get_finance_data(yr = "all", geo = "all", dataset_type = "skinny")
```

## `NA` is not zero

Some F-33 items were originally coded `-1` or `-2` to indicate missing values; `edfinr` recodes these to `NA` during processing. An `NA` therefore means the item was **not reported**, not that the district spent or received nothing. Dropping `NA`s silently (for example, with `na.rm = TRUE` in a ranking) keeps incompletely-reporting districts in your analysis with partial values, which can be worse than excluding them; decide explicitly.

Under version 0.2.0 (and beyond), `edfinr` also detects a second F-33 missingness convention: some items are **zero-filled** when missing, with a companion data-item flag (`FL_* = "M"`) marking them unreported. The data processing proceture converts these flagged zero-fills to `NA` for the COVID (`exp_covid_*`), capital-detail, debt, fund-balance, CE fund-type, and expenditure-detail items. The largest effect is on the COVID columns: all of New York (including NYC) in every year and roughly a third to half of California districts from FY2021 onward never reported them, and those districts previously appeared as spending exactly $0. Genuine reported zeros are preserved. Missingness also **propagates** through derived measures rather than being treated as zero. 

## The state-revenue adjustment and `c11_spike_flag`

Following the EdBuild methodology, `edfinr` nets capital and debt-related state aid (F-33 item C11) out of `rev_state`, so that `rev_state` approximates operating support. The pre-adjustment values are preserved in `rev_state_unadj` and `rev_state_unadj_pp`, and the netted-out C11 amount itself ships as `rev_state_cap_debt` (in both datasets), so the adjustment can be reconstructed directly: `rev_state_unadj - rev_state_cap_debt - rev_state` equals the state share of the other-system-payment adjustment. Because `rev_state_cap_debt` feeds the adjustment arithmetic, it is zero-filled, not `NA`, where districts did not report.

In a small number of district-years, the C11 adjustment produces an anomalous spike -- typically when a large one-time capital grant flows through -- and `c11_spike_flag` marks them:

```{r spike-counts}
us |>
  group_by(year) |>
  summarize(flagged = sum(c11_spike_flag, na.rm = TRUE), .groups = "drop")

# which states account for the most flagged district-years?
us |>
  filter(c11_spike_flag) |>
  count(state, sort = TRUE) |>
  head(10)
```

For a flagged district, the adjusted state-revenue series can dip or jump in ways the unadjusted series does not. Comparing the two makes the artifact obvious. Here we pick the largest district with multiple flagged years:

```{r spike-example, fig.height = 4.5, fig.alt = "Line chart comparing adjusted and unadjusted state revenue per-pupil for one district from 2012 to 2023, with triangle markers on the years flagged by c11_spike_flag where the two series diverge sharply."}
example_id <- us |>
  filter(c11_spike_flag) |>
  count(ncesid, wt = enroll, sort = TRUE) |>
  slice(1) |>
  pull(ncesid)

us |>
  filter(ncesid == example_id) |>
  select(year, dist_name, c11_spike_flag, rev_state_pp, rev_state_unadj_pp) |>
  pivot_longer(
    cols = c(rev_state_pp, rev_state_unadj_pp),
    names_to = "series", values_to = "value"
  ) |>
  mutate(series = ifelse(
    series == "rev_state_pp", "Adjusted (rev_state_pp)", "Unadjusted (rev_state_unadj_pp)"
  )) |>
  ggplot(aes(x = year, y = value, color = series)) +
  geom_line() +
  geom_point(aes(shape = c11_spike_flag), size = 2.5) +
  scale_x_continuous(breaks = seq(2013, 2023, 2)) +
  scale_y_continuous(labels = scales::label_dollar()) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17)) +
  labs(
    title = "Adjusted vs. Unadjusted State Revenue Per-Pupil",
    subtitle = "Triangles mark years where c11_spike_flag is TRUE",
    x = "Year", y = "State Revenue Per-Pupil",
    color = NULL, shape = "Flagged"
  ) +
  theme_minimal()
```

When analyzing state revenue over time, check `c11_spike_flag` for your districts of interest, inspect `rev_state_cap_debt` to see how large the adjustment actually is, and consider using `rev_state_unadj_pp` (or excluding flagged years) where the adjustment dominates the series.

## Pass-through districts and `osp_pct`

Some districts pass a large share of the revenue they receive on to other school systems (charter schools, other LEAs, private placements). `edfinr` proportionally subtracts these payments from revenues, and `osp_pct` reports the share of unadjusted total revenue paid to other systems, so you can see how much the adjustment matters for any district:

```{r osp}
us_2023 <- us |> filter(year == 2023)

# how common are large pass-through shares?
us_2023 |>
  summarize(
    n = sum(!is.na(osp_pct)),
    over_10_pct = sum(osp_pct > 0.10, na.rm = TRUE),
    over_25_pct = sum(osp_pct > 0.25, na.rm = TRUE)
  )

# the largest pass-through districts, among districts with 1,000+ students
us_2023 |>
  filter(enroll >= 1000) |>
  arrange(desc(osp_pct)) |>
  select(dist_name, state, enroll, osp_pct, rev_total_pp) |>
  head(10)
```

For high-`osp_pct` districts, per-pupil figures describe the students the district serves directly, after the pass-through dollars are removed. If you are more interested in every dollars a district handles, use the `rev_*_unadj` variables instead.

## Panel composition: districts enter and exit

The universe of districts in `edfinr` is not constant across years. The largest single discontinuity is in California: beginning in school year 2018-19, a wave of California charter schools switched to independent reporting and were assigned their own NCES LEA IDs for the first time. Once in the NCES LEA universe, those new charter-LEAs automatically show up in the F-33 finance survey. The result is a jump in California district counts from 2019 onward that reflects reporting structure, not new schools:

```{r panel-plot, fig.height = 4.5, fig.alt = "Two-panel line chart of district counts by year for California and all other states; the California count jumps sharply in 2019 when charter schools became separately reported LEAs."}
us |>
  mutate(group = ifelse(state == "CA", "California", "All other states")) |>
  count(group, year) |>
  ggplot(aes(x = year, y = n)) +
  geom_line() +
  geom_point(size = 1) +
  facet_wrap(~group, scales = "free_y") +
  scale_x_continuous(breaks = seq(2013, 2023, 2)) +
  labs(
    title = "Districts in the edfinr Panel by Year",
    subtitle = "The 2019 jump in California reflects charter schools becoming separate LEAs",
    x = "Year", y = "Districts"
  ) +
  theme_minimal()
```

For longitudinal work, consider a **consistent sample**: keep only districts present in every year of your window.

## Variable availability by year

Not every variable spans the full 2012-2023 panel. The data dictionary's `first_yr_avail` column records when each variable enters the `edfinr` dataset:

```{r availability}
list_variables("full") |>
  filter(first_yr_avail != "2012") |>
  select(name, first_yr_avail, description) |>
  arrange(first_yr_avail, name) |>
  knitr::kable()
```

A multi-year average or trend that crosses one of these boundaries will mix real change with a coverage change. 

## Urbanicity: condensed vs. raw locale codes

The `urbanicity` factor condenses the NCES urban-centric locale codes into four categories (City, Suburb, Town, Rural). The raw 12-category codes are also available as `urbanicity_raw` (numeric) and `urbanicity_raw_cat` (labeled factor):

```{r urbanicity}
us_2023 |>
  count(urbanicity_raw_cat, urbanicity) |>
  arrange(urbanicity, urbanicity_raw_cat)
```

The condensed categories are right for most summaries, but the raw codes can be helpful when the distinction *within* a category matters. For example, "Rural: Fringe" districts sit at metro edges and often look more like suburbs than like "Rural: Remote" districts.

## Comparability checklist

Before publishing numbers built on `edfinr` data:

1. **Check missingness** for your variables and years; decide explicitly how to treat `NA` district-years rather than relying on `na.rm = TRUE`.
2. **Check `first_yr_avail`** for every variable in a multi-year analysis.
3. **Check `c11_spike_flag`** (and compare against `rev_state_unadj_pp`) when state revenue is central to the analysis.
4. **Check `osp_pct`** when high-pass-through districts could distort per-pupil comparisons.
5. **Use a consistent sample** (or justify not doing so) for trend analyses, especially any that include California.
6. **Use CPI adjustment** for any multi-year dollar comparison (see the "CPI Adjustments" vignette), and remember debt and fund-balance stocks stay nominal.

## See also

- The "Data Sources and Methodology" vignette for the processing pipeline, exclusion rules, and the F-33 crosswalk.
- The "CWIFT" article for imputation flags on the wage index.
- The "Capital and Facilities" article for lumpiness in capital outlay.
