---
title: "Mapping School Finance Data"
---

```{r, include = FALSE}
# this article needs sf and tigris (declared in Config/Needs/website);
# skip evaluation gracefully when they are not installed
has_geo <- requireNamespace("sf", quietly = TRUE) &&
  requireNamespace("tigris", quietly = TRUE)
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  eval = has_geo
)
```

## Introduction

Maps are often the most effective way to communicate school finance patterns:
funding disparities, labor-cost geography, and property-wealth gradients all
have strong spatial structure. `edfinr` does not ship geometries, but its
`ncesid` matches the GEOID used by Census Bureau school district boundary
files -- the same identifier the package itself uses to join ACS data -- so
joining finance data to shapes from the [tigris](https://github.com/walkerke/tigris)
package is a one-liner.

```{r setup, include = FALSE}
library(edfinr)
library(dplyr)
library(ggplot2)
library(sf)
library(tigris)
options(tigris_use_cache = TRUE)
```

```{r, eval = FALSE}
library(edfinr)
library(dplyr)
library(ggplot2)
library(sf)
library(tigris)

# cache downloaded shapefiles across sessions
options(tigris_use_cache = TRUE)
```

## Before you map: district geography caveats

1. **Three overlapping layers.** Census publishes *unified*, *elementary*, and
   *secondary* school district boundaries. In states where elementary and
   secondary districts coexist (Illinois, California, and others), the layers
   overlap, and mapping only the unified layer silently drops districts.
   `tigris::school_districts()` takes a `type` argument; check your state's
   structure before assuming `"unified"` covers it.
2. **Boundaries change.** Use a geometry `year` close to your data year, since
   district consolidations and boundary changes accumulate.
3. **Cartographic boundary files.** `cb = TRUE` returns generalized boundaries
   that are smaller and draw faster; use the default TIGER/Line files only
   when you need legal-boundary precision.
4. **Not every district joins.** Some `edfinr` districts (certain charter
   LEAs, for example) have no mapped boundary, and some mapped areas have no
   finance record. Count the misses after joining rather than assuming
   completeness.

## Joining finance data to boundaries

Ohio's districts are all unified, which keeps the example simple. Pull the
finance data and the boundaries, then join on `GEOID = ncesid`:

```{r join}
oh_2023 <- get_finance_data(yr = "2023", geo = "OH")

oh_shapes <- school_districts(state = "OH", year = 2023, cb = TRUE)

oh_map <- oh_shapes |>
  left_join(oh_2023, by = c("GEOID" = "ncesid"))

# diagnostics: how many shapes have no finance record, and vice versa?
sum(is.na(oh_map$rev_total_pp))
nrow(anti_join(oh_2023, sf::st_drop_geometry(oh_shapes), by = c("ncesid" = "GEOID")))
```

## Mapping per-pupil revenue

With an `sf` object, `ggplot2::geom_sf()` handles the rest. Binned scales
usually communicate funding levels better than continuous gradients, because
readers can attach a dollar range to each color:

```{r revenue-map, fig.height = 6.5}
ggplot(oh_map) +
  geom_sf(aes(fill = rev_total_pp), color = "white", linewidth = 0.05) +
  scale_fill_viridis_b(
    n.breaks = 6,
    labels = scales::label_dollar()
  ) +
  labs(
    title = "Total Revenue Per-Pupil in Ohio Districts, SY2022-23",
    fill = "Revenue\nPer-Pupil"
  ) +
  theme_void()
```

## Mapping labor costs with CWIFT

Any `edfinr` variable maps the same way. The CWIFT teacher-wage index makes
the metro/rural labor-cost geography visible (see the "CWIFT" article for what
the index does and does not measure):

```{r cwift-map, fig.height = 6.5}
ggplot(oh_map) +
  geom_sf(aes(fill = cwift_est), color = "white", linewidth = 0.05) +
  scale_fill_viridis_c(option = "magma") +
  labs(
    title = "Comparable Wage Index for Teachers in Ohio, FY2023",
    subtitle = "FY2023 values are carried forward from FY2022",
    fill = "CWIFT"
  ) +
  theme_void()
```

## Going further

- **Multi-state maps**: `school_districts()` accepts a vector of states;
  filter `get_finance_data(geo = ...)` to match.
- **Coastal states**: `tigris::erase_water()` improves shoreline maps.
- **ACS overlays**: the [tidycensus](https://walker-data.com/tidycensus/)
  package retrieves ACS variables with geometry included
  (`geometry = TRUE`); its GEOIDs join to `ncesid` the same way, which is
  convenient when you need ACS measures beyond those bundled in `edfinr`.
- **Interactive maps**: `sf` objects work directly with `mapview` and
  `leaflet` for exploratory work.

## See also

- The "CWIFT" article for interpreting the wage index.
- The "Community and Economic Context" article for the ACS and SAIPE measures worth mapping.
- The "Data Quality and Comparability" article for panel-composition caveats that affect join completeness.
